---
title: "PSTAT 105 - Assignment 2"
author: "Roshan Mehta"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1 - Basketball B-Day Data
A well-known analysis in Malcolm Gladwell’s book Outlier argues that the best hockey players are more likely to be born earlier in the year presumably because this gives them advantages in the youth hockey leagues. We are interested in checking whether there is a similar effect in basketball.

The data set Basketball_Ref_BDays.txt contains information for a large sample of professional basketball players listed on the http://www.basketball-reference.com website. Use the `table` or `dplyr::count` function  to  calculate  how  many  players  were  born  in  each  month.   Draw  an appropriate plot.
```{r}
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
```
 

a)
```{r}
#setwd("/Users/roshanmehta/Downloads/PSTAT/PSTAT 105/pstat105-nonparametric-methods/data_files")

## Reading in the entire data file with some special
## formatting to interpret the birthday dates correctly
BasketBDays <- read_csv("https://github.com/roshanmehta12/pstat105-nonparametric-methods/raw/main/data_files/Basketball_Reference_BDays.txt", 
                        col_types = cols(`Birth Date`= col_date(format = "%B %d %Y")))

## There are some players without birthdays, we should drop them
sum(is.na(BasketBDays$`Birth Date`))
BasketBDays <- BasketBDays %>% drop_na(`Birth Date`)

## format allows us to extract the month of their birth
BasketBDays$Month <- format(BasketBDays$`Birth Date`,format="%m")
BasketBDays$Year <- format(BasketBDays$`Birth Date`,format="%Y")
BasketBDays$Day <- format(BasketBDays$`Birth Date`,format="%d")

## count is a tidyverse function which tabulates the outcomes
BBall.table <- BasketBDays %>% count(Month)
```
Here is the data tabulated by month.
$$
\begin{array}{ccccccc}
\hline \text { Month } & \text { Jan } & \text { Feb } & \text { Mar } & \text { Apr } & \text { May } & \text { Jun } \\
\text { Count } & 430 & 418 & 445 & 366 & 402 & 420 \\
\hline \text { Month } & \text { Jul } & \text { Aug } & \text { Sep } & \text { Oct } & \text { Nov } & \text { Dec } \\
\text { Count } & 462 & 417 & 443 & 431 & 397 & 387 \\
\hline
\end{array}
$$
We can represent this data in a bar graph as in figure 1.  The horizontal line represents the level if all the months had an equal number of births.

```{r}
ggplot(BBall.table, aes(x=Month, y=n)) + 
  geom_col(fill="violet") + 
  labs(y="Count") + 
  geom_hline(yintercept = mean(BBall.table$n), size=1.4, color="red")
```
Figure 1: The number of basketball players born in each month.  The red line indicates the average over all months.


b) Perform a $\chi^2$ test to see if the players are equally likely to be born in any month.
```{r}
expected <- sum(BBall.table$n[1:12])/12
X2 <- sum((BBall.table$n[1:12]-expected)^2/expected)
X2
1 - pchisq(X2,df=11)

# or
chisq.test(BBall.table$n[1:12])
```
Therefore, there is not a significant difference between the observed birthdays and just an even distribution across the months.

c) In order to focus our attention on modern players, repeat this analysis with only those players
that were born after 1/1/1955. (also use this smaller data set for the following questions.)
```{r}
BasketBDays$Year <- format(BasketBDays$`Birth Date`, format="%Y")
BBTableMnths.modern  <- filter(BasketBDays, Year >= 1955) %>% count(Month)
BBTableMnths.modern
```

Then we can use this tabulation to produce the barplot in figure 2.
The $\chi^2$ test is produced just as it was before. 
```{r}
expct.n <- mean(BBTableMnths.modern$n)
X2 <- sum((BBTableMnths.modern$n-expct.n)^2/expct.n)
X2
1 - pchisq(X2, df=11)
```
This  statistic  shows  that  there  is  still  no  evidence  that  players  birthdays  are  clustered  in  any particular months.

```{r}
ggplot(BBTableMnths.modern, aes(x=Month, y=n)) + 
  geom_col(fill="violet") + 
  labs(y="Count") + 
  geom_hline(yintercept = mean(BBTableMnths.modern$n), 
             size=1.4, 
             color="red")
```
Figure 2:   The number of players born in each month for players born after 1954.  There does appear to be fewer than average birthdays in April, November and December.


d) To be more careful, we should realize that more people are probably born in January than February just because there are more days in January. Perform a $\chi^2$ test where the null hypothesis is that the probability of each month is proportional to the average number of days in that month.

To be more careful, we should realize that more people are probably born in January than February just because there are more days in January. As a result, we should be testing the null hypothesis
$$
\begin{array}{lrrrrrr}
\hline \text { Month } & \text { Jan } & \text { Feb } & \text { Mar } & \text { Apr } & \text { May } & \text { Jun } \\
\text { Days } & 31 & 28.25 & 31 & 30 & 31 & 30 \\
\text { Prob. } & 0.085 & 0.077 & 0.085 & 0.082 & 0.085 & 0.082 \\
\hline \text { Month } & \text { Jul } & \text { Aug } & \text { Sep } & \text { Oct } & \text { Nov } & \text { Dec } \\
\text { Days } & 31 & 31 & 30 & 31 & 30 & 31 \\
\text { Prob. } & 0.085 & 0.085 & 0.082 & 0.085 & 0.082 & 0.085 \\
\hline
\end{array}
$$
Note how we have dealt with the issue of Leap Years (the players were born in months over many years so we expect about 1 in 4 years to include leap years.  There were only 3 players in the modern data set, Chucky Brown, Vonteego Cummings, and Tyrese Haliburton born on Feb. 29.)

```{r}
LeapYears <-  filter(BasketBDays, Month == "02" & Day == "29")
LeapYears$Player
```
Here is the calculation of the $\chi^2$ statistic
```{r}
p <- c(31,28.25,31,30,31,30,31,31,30,31,30,31)/365.25
chisq.test(BBTableMnths.modern$n, p=p)
```
The data fits this new distribution worse than before, but there is not a statistically significant difference from the proportion of days.  We should accept the null hypothesis in this case as well.

e) Going even further, it seems that some months generally are favored over others for having babies (summer births are more likely). We should probably compare our basketball player data to the following probabilities from the CDC.
$$
\begin{array}{ccccccc}
\hline \text { Month } & \text { Jan } & \text { Feb } & \text { Mar } & \text { Apr } & \text { May } & \text { Jun } \\
\hline \text { Prob. } & 0.0815 & 0.0752 & 0.0837 & 0.0816 & 0.0860 & 0.0813 \\
\hline \hline \text { Month } & \text { Jul } & \text { Aug } & \text { Sep } & \text { Oct } & \text { Nov } & \text { Dec } \\
\hline \text { Prob. } & 0.0883 & 0.0892 & 0.0866 & 0.0849 & 0.0787 & 0.0830 \\
\hline
\end{array}
$$
This leads us to another version of the $\chi^2$ test.
```{r}
n <- sum(BBTableMnths.modern$n)
pc = c(0.0815 , 0.0752 , 0.0837 , 0.0816 , 0.0859 , 0.0813,0.0883 , 0.0892 , 0.0866 , 0.0849 , 0.0787 , 0.0831)
chisq.test(BBTableMnths.modern$n,p=pc)
```
This test is once again further from our data, but not to a statistically significant degree.

f) Interpret your results. Is there significant evidence at an $\alpha = 0.05$ level that professional basketball players are born earlier in the year than the normal population?

This data does not show a statistically significant tendency for professional basketball players to be born earlier in the year.

```{r}
expectc <- n*pc
ggplot(BBTableMnths.modern, aes(x=Month, y=n)) + 
  geom_col(fill="violet") + 
  labs(y="Count") + 
  annotate(geom="point", x=1:12, y = expectc)
```
Figure 3:    A comparison of the counts for each month versus the expected values from the CDC numbers represented by the dots.


Figure 3 shows a comparison from this last test between which months the players were born and which months the CDC reported as the general population proportions.From this plot, it seems that the main difference is that a lot more players are born in February and fewer are born in August and December.  This lends some evidence to our theory that children born earlier in the year are more likely to become professional athletes.  Though, the difference is not large enough to be significant.The first two tests are less significant probably because the effect of an advantage for players born earlier  in  the  year  is  somewhat  counteracted  by  families  having  fewer  children  in  those  winter months.


# 2 - Selling times data














































 